---
title: An R Markdown document converted from "C:/Users/vdtoo/SynologyDrive/Work/teaching/2024/Development
  omics in oncology COIL/repo/R_exercises/02 Wednesday/4. Reading in data.ipynb"
output: html_document
---

# Reading data

Usually, you don't want to enter in the data into R manually, but you will get the data from an external source.  
Often this will be a text file, written out by some analysis program. Sometimes it is data that is uploaded to a journal in excel format.

There are several methods to read data, we'll be using packages from some libraries.

## Simple text

R can read text files (such as csv or tsv files). I have downloaded one file from https://datacarpentry.org/semester-biology/materials/datasets/.

The most commonly used text formats are the following:

- csv, "comma separated values". This is a file format that has lines \(rows\) with values, with commas in between. Usually the first line contains the column headers.
  - Some countries \(notably The Netherlands\) use a comma instead of a dot. csv files exported from Excel usually replace the 'comma between values' with a semicolon \( ; \).
- tsv, "tab separated values".
  - Many times tsv files are stored with the 'txt' file extension. Always look inside of the file with a text editor if you are not sure wha  
    t the delimiter is.
  - If you need to write down the character "tab" in a command, you can use "_\\t_" \(similar to _\\n_ which means newline\)

#### Method 1: 'base' R file reading

R has different read methods. The help page for many of them is combined, you can find it by getting help for `'read.delim`'.

```{r}
#find information about read.delim
? read.delim
```

#### :ballot_box_with_check: Read in the avian_ssd_jan07.txt file

Put the result in the variable with name 'birds'. Have a look at the file, and decide what type of delimited file we're looking at, and choose a reading function.

Have a look at the contents of 'birds', does some number appear over and over again in unexpected places?  


```{r}
birds <- read.delim("avian_ssd_jan07.txt")
```

```{r}
birds
```

#### :ballot_box_with_check: Small adjustments for reading in the data

It turns out the number you found, is actually used here for as "not a number", or "not valid" or "missing" value. 

Because the people who created the file did not use NA in the file, R just shows the number. Even though bit of a strange thing to do, but we'll have to deal with it.

To read in the file in a better way, we can tell R to read the number you found as if it is NA. The trick to do this can be found in the documentation. It is a special argument for the function that tells R to interpret certain numbers as 'NA'.  
Read in the data with this extra argument for the read function. Note, there are actually two \(slightly different\) representations of the same number, you will need to tell R what variants exist.  


```{r}
# Read the bird data in, with NA's
bird <- read.delim("avian_ssd_jan07.txt", na=c("-999", "-999.0"))
```

```{r}
bird
```

## Reading text with the Tidyverse functions

If you load in the 'tidyverse' library, you will get access to some new (improved) functions for reading in text. These functions work the same way as the 'base' R functions, with the exception that they have been made faster, and also that the result that they produce is a Tibble, if possible.

#### :ballot_box_with_check: Load the 'tidyverse' library, and use read_delim (see the difference in the naming?) to read in the same file.


Name the variable 'birds2'.

```{r}
# Read in the tidyverse library
library("tidyverse")
```

```{r}
# don't forget 'str'
birds2 <- read_delim("avian_ssd_jan07.txt")

class(birds2)
```

## Excel

Load in the 'readxl' library. Read the help file about how to read in a file.  
There is an Excel file for an exercise that we are going to use in another exercise. Let's use that to load in the data.  
To get to the file, use the following character string: "../../data/CELL-179-suppl1.xlsx"

Read the data and check if everything went ok. If not, check the excel file and try to fix it!


```{r}
# Load the readxl library
library(readxl)
```

```{r}
# Read the file in "../../data/CELL-179-suppl1.xlsx"
cell <- read_excel("../../data/CELL-179-suppl1.xlsx", skip=1)
cell
```

# Use Cheat Sheets! (tidyverse.org)

